---
title: "Sample Quarto Document as Slides"
author: "John Curtin"
format:
  revealjs:
    scrollable: true
    css: slides.css
    slide-number: true
    show-slide-number: all
editor_options: 
  chunk_output_type: console
---

This is a sample Quarto document that will be rendered as slides.

All the same as before but added a few slides at the end without headings.


## Introduction

This is a simple notebook that demonstrates the use of the `quarto` package.

We can use section headings of different levels using `#`, `##`, `###`, etc.

There are lots of other formatting options for Markdown text, including **bold**, *italics*, and `code`.

## Lists

We can also make lists

- item 1
- item 2
- item 3

## Code Chunks
And we can include code chunks

```{r}
#| warning: false
#| message: false

library(tidyverse)
```

# Plots

And we can include plots

```{r}
ggplot(data = mtcars) +
  geom_point(aes(x = wt, y = mpg))
```

## Tables

And we can include tables

```{r}
head(mtcars)
```

-----

Here is a slide that didnt have a header.  

The new slide was denoted with `-----`
