---
title: "Sample Quarto Document"
author: "John Curtin"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
editor_options: 
  chunk_output_type: console
---

The YAML above includes info for formatting html

 - Includes a toc with depth of 2
 - Embeds resources in the html file (a single file with all resources)
 - Default is now to render to html with `quarto render 3_quarto.qmd`
 - Can still render to pdf with: `quarto render 3_quarto.qmd --to pdf`
 
## Introduction

This is a simple notebook that demonstrates the use of the `quarto` package.

We can use section headings of different levels using `#`, `##`, `###`, etc.

There are lots of other formatting options for Markdown text, including **bold**, *italics*, and `code`.

## Lists

We can also make lists

- item 1
- item 2
- item 3

## Code Chunks
And we can include code chunks

```{r}
#| warning: false
#| message: false

library(tidyverse)
```

# Plots

And we can include plots

```{r}
ggplot(data = mtcars) +
  geom_point(aes(x = wt, y = mpg))
```

## Tables

And we can include tables

```{r}
head(mtcars)
```
