---
title: auROCs for Lagged models 
author: John Curtin 
editor_options: 
  chunk_output_type: console
---


NOTE THESE ARE TEMP FIGS BASD ON  INNER LOOP PERFORMANCE.  NEED TO UPDATE TO OUTER LOOP AND ALSO TO FILES THAT START WITH POSTERIOR.

## Set up
```{r}
# handle conflicts
options(conflicts.policy = "depends.ok")

library(tidyverse)
library(tidyposterior)
theme_set(theme_classic()) 

# Paths
devtools::source_url("https://github.com/jjcurtin/lab_support/blob/main/format_path.R?raw=true",
                     sha1 = "a58e57da996d1b70bb9a5b58241325d6fd78890f")
# source("../../lab_support/format_path.R")
path_lag <- format_path(str_c("studydata/risk/models/lag"))
```

## Get/Process data

This commented out code chunk makes a tmp file of 300 inner folds for each of 5 models and saves on server.   This will change once we have outer loop performance.
```{r}
#inner_lag <- read_csv(file.path(path_lag, 
#                                 "inner_metrics_1week_0_v2_nested_main.csv"),
#                       col_types = cols()) |> 
#  mutate(lag = 0) |> 
#  bind_rows(read_csv(file.path(path_lag, 
#                                 "inner_metrics_1week_24_v2_nested_main.csv"),
#                       col_types = cols()) |> 
#  mutate(lag = 24)) |> 
#  bind_rows(read_csv(file.path(path_lag, 
#                                 "inner_metrics_1week_72_v2_nested_main.csv"),
#                       col_types = cols()) |> 
#  mutate(lag = 72)) |> 
#  bind_rows(read_csv(file.path(path_lag, 
#                                   "inner_metrics_1week_168_v2_nested_main.csv"),
#                         col_types = cols()) |> 
#  mutate(lag = 168)) |> 
#  bind_rows(read_csv(file.path(path_lag, 
#                                   "inner_metrics_1week_336_v2_nested_main.csv"),
#                         col_types = cols()) |> 
#  mutate(lag = 336)) |> 
#  glimpse()
#
#inner_best <- inner_lag |> 
#  group_by(algorithm, feature_set, hp1, hp2, hp3, resample, lag)  |>  
#    summarize(across(c(accuracy, bal_accuracy, roc_auc, sens, spec, ppv, npv),
#                     median),
#              n_jobs = n(), .groups = "drop") |> 
#  group_by(lag) |> 
#  arrange(desc(roc_auc)) |> 
#  slice(1) |> 
#  rename(med_roc_auc = roc_auc) |> 
#  select(lag, algorithm, hp1, hp2, hp3, resample, med_roc_auc, n_jobs)
#
#inner_lag <- inner_lag |> 
#  full_join(inner_best, by = c("lag", "algorithm", "hp1", "hp2", "hp3", "resample")) |> 
#  filter(!is.na(med_roc_auc))


inner_lag <- read_csv(here::here(path_lag, "tmp_300folds.csv"))
```

NOTE: THIS IS ALSO NOT QUITE RIGHT. NEED TO BE MODELED ON FOLDS AND REPEATS
```{r}
#set.seed(101)
#pp <- inner_lag |> 
#  select(outer_split_num, inner_split_num, lag, roc_auc) |> 
#  pivot_wider(names_from = lag, values_from = roc_auc) |> 
#  rename(id = outer_split_num,
#         id2 = inner_split_num) |> 
#  perf_mod(formula = statistic ~ model + (1 | id2/id),
#         transform = tidyposterior::logit_trans,  # for skewed & bounded AUC
#         adapt_delta = .99,
#         family = gaussian, 
#)  

#pp_tidy <- pp |> 
#  tidy(seed = 123) 

#q = c(.025, .5, .975)
#ci <- pp_tidy |> 
#  group_by(model) |> 
#  summarize(pp_median = quantile(posterior, probs = q[2]),
#            pp_lower = quantile(posterior, probs = q[1]), 
#            pp_upper = quantile(posterior, probs = q[3])) |> 
#   mutate(model = factor(model, levels = c(0, 24, 72, 168, 336),
#                        labels = c("0 lag", "24 lag", "72 lag", "168 lag", "336 lag"))) |> 
#   arrange(model)

ci <- read_csv(here::here(path_lag, "tmp_ci.csv")) |> 
  mutate(model = case_match(model,
                            "0 lag" ~ "No lag",
                            "24 lag" ~ "1 day lag",
                            "72 lag" ~ "3 day lag",
                            "168 lag" ~ "1 week lag",
                            "336 lag" ~ "2 week lag")) |> 
  mutate(model = fct(model, 
                     levels = c("2 week lag", "1 week lag", "3 day lag", 
                                "1 day lag", "No lag"))) |> 
  glimpse() 
```


```{r}
#| label: fig-lag_auroc 
#| fig-cap: auROCs by Prediction Window Lag
#| fig-height: 6
#| fig-width: 6

ci |>
  ggplot(aes(x = model)) +
  geom_point(aes(y = pp_median), size = 3) +
  geom_errorbar(aes(ymin = pp_lower, ymax = pp_upper),
    width = .2,
    position = position_dodge(.9)
  ) +
  coord_flip() +
  ylab("auROC") +
  xlab("Prediction Window Lag") +
  theme(legend.position = "none") +
  scale_y_continuous(
    breaks = seq(0.8, 1, 0.05),
    limits = c(0.8, 1)
  ) +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16))
```